---
title: "Stability of political stringency - A multivariate approach"
author: Romain Frelat
date: today
format: 
  html:
    toc: true
execute:
  echo: false
  warning: false
---

## Objective

The idea is to use a *multivariate* length and see :  

- what are the main political stringency "axis" : i.e. differences among communes in political stringency indicators
- are the main axis of variations in political stringency indicators similar if calculated per regions?


```{r}
# Load home made functions
devtools::load_all()

ref_folder <- here::here("data", "derived-data", "ref")
ind_folder <- here::here("data", "derived-data", "indicators_csv")

reg <- terra::vect(file.path(ref_folder, "gadm41_FRA_1.json"))
dep <- terra::vect(file.path(ref_folder, "gadm41_FRA_2.json"))
shp <- terra::vect(file.path(ref_folder, "commune_simple0005_4326.gpkg"))

# meta <- readxl::read_xlsx(
#   here::here("data", "derived-data", "megatrends_metadata.xlsx")
# )
# ps_crit <- grepl("stringency", meta$Megatrend)
# ind_crit <- meta$Indicator == "yes"
# ps <- meta[ps_crit & ind_crit, ]
var <- c(
  "BIO_AREA_PCT_2024",
  "PROT_AREA_PCT_2026",
  "PESTICIDE_EXPOSURE_2022",
  "CROP_DIV_SHANNON_2023",
  "FALLOW_AREA_PCT_2023",
  "CROPROT_INV_SIM_1523"
)
df <- read.csv(file.path(ind_folder, "Dataset_commune.csv"))

ind <- df[, var]

# remove NAs
keep <- complete.cases(ind)
ind <- ind[keep, ]
df <- df[keep, ]

shp <- shp[match(df$INSEE_COM, shp$INSEE_COM)]
```

### Background information

#### Variables

```{r}
#| label: metadata
meta <- readxl::read_xlsx(
  here::here("data", "derived-data", "megatrends_metadata.xlsx")
)
colC <- c(
  "Name",
  "Description",
  "Year",
  "Unit",
  "Source"
)
ps <- meta[meta$Name %in% var, colC]
knitr::kable(ps)
```

#### Regions

| Code | Region                     |
|------|----------------------------|
| 11   | ILE DE FRANCE              |
| 24   | CENTRE VAL DE LOIRE        |
| 27   | BOURGOGNE FRANCHE COMTE    |
| 28   | NORMANDIE                  |
| 32   | HAUTS DE FRANCE            |
| 44   | GRAND EST                  |
| 52   | PAYS DE LA LOIRE           |
| 53   | BRETAGNE                   |
| 75   | NOUVELLE AQUITAINE         |
| 76   | OCCITANIE                  |
| 84   | AUVERGNE RHONE ALPES       |
| 93   | PROVENCE ALPES COTE D AZUR |
| 94   | CORSE                      |



::: {.panel-tabset}
## scaled variables

```{r}
boxplot(scale(ind), las = 2)
```

## square root transformed

```{r}
boxplot(scale(apply(ind, 2, sqrt)), las = 2)
```

::: 


## France scale

```{r}
library(ade4)

pca1 <- ade4::dudi.pca(
  apply(ind, 2, sqrt),
  scannf = FALSE,
  nf = 2
)

ade4::s.class(pca1$li, as.factor(df$INSEE_REG), col = rainbow(13))
ade4::s.arrow(pca1$co * 5, add.plot = TRUE)
```


::: {.panel-tabset}
## PC1
```{r}
minmax1 <- max(abs(pca1$li[, 1]))
shp$PC1 <- pca1$li[, 1]

# PC1
terra::plot(
  shp,
  "PC1",
  main = "PC1",
  border = NA,
  breaks = seq(-minmax1, minmax1, length.out = 7)
)
terra::plot(reg, add = TRUE, lwd = 2)
```

## PC2

```{r}
minmax2 <- max(abs(pca1$li[, 2]))
shp$PC2 <- pca1$li[, 2]
# PC2
terra::plot(
  shp,
  "PC2",
  main = "PC2",
  border = NA,
  breaks = seq(-minmax2, minmax2, length.out = 7)
)
terra::plot(reg, add = TRUE, lwd = 2)
```

::: 


## Regional effect


```{r}
varPC1 <- pca1$co[, 1]
varPC2 <- pca1$co[, 2]

# keep information on PC1
lab1 <- paste0("France_PC1_", round(pca1$eig[1] / sum(pca1$eig) * 100), "P")
PC1 <- data.frame(varPC1)
colnames(PC1) <- lab1
rownames(PC1) <- row.names(pca1$co)

lab2 <- paste0("France_PC2_", round(pca1$eig[2] / sum(pca1$eig) * 100), "P")
PC2 <- data.frame(varPC2)
colnames(PC2) <- lab2
rownames(PC2) <- row.names(pca1$co)

reg <- sort(unique(df$INSEE_REG))
for (i in reg) {
  df_reg <- ind[df$INSEE_REG == i, ]
  pca_i <- ade4::dudi.pca(
    apply(df_reg, 2, sqrt),
    scannf = FALSE,
    nf = 2
  )
  # so far, PC1 = PC1; PC2 = PC2
  # only change the sign
  if (cor(pca_i$co[, 1], varPC1) < 0) {
    pca_i$co[, 1] <- -pca_i$co[, 1]
    pca_i$li[, 1] <- -pca_i$li[, 1]
  }
  if (cor(pca_i$co[, 2], varPC2) < 0) {
    pca_i$co[, 2] <- -pca_i$co[, 2]
    pca_i$li[, 2] <- -pca_i$li[, 2]
  }

  # PC1
  #fmt: skip
  lab1_i <- paste0("R", i,"_PC1_",
    round(pca_i$eig[1] / sum(pca_i$eig) * 100), "P")
  PC1 <- cbind(PC1, pca_i$co[, 1])
  names(PC1)[ncol(PC1)] <- lab1_i

  # PC2
  #fmt: skip
  lab2_i <- paste0("R", i,"_PC2_",
    round(pca_i$eig[2] / sum(pca_i$eig) * 100), "P")
  PC2 <- cbind(PC2, pca_i$co[, 2])
  names(PC2)[ncol(PC2)] <- lab2_i
}
```

::: {.panel-tabset}
## PC1

```{r}
ord1 <- order(PC1[, 1])

dotchart(
  PC1[ord1, 1],
  pch = 16,
  pt.cex = 2,
  xlim = range(PC1),
  labels = row.names(PC1)[ord1],
  main = "PC1"
)
for (i in 2:ncol(PC1)) {
  points(
    PC1[ord1, i],
    1:nrow(PC1),
    pch = 16,
    col = rainbow(ncol(PC1))[i],
    add = TRUE
  )
}
abline(v = 0, lty = 2)
legend(
  "topleft",
  legend = colnames(PC1),
  pch = 16,
  col = c("black", rainbow(ncol(PC1))),
  pt.cex = c(2, rep(1, ncol(PC1))),
  bty = "n",
  cex = 0.7
)
```

## PC2

```{r}
ord2 <- order(PC2[, 1])

dotchart(
  PC2[ord2, 1],
  pch = 16,
  pt.cex = 2,
  xlim = range(PC2),
  labels = row.names(PC2)[ord2],
  main = "PC2"
)
for (i in 2:ncol(PC2)) {
  points(
    PC2[ord2, i],
    1:nrow(PC2),
    pch = 16,
    col = rainbow(ncol(PC2))[i],
    add = TRUE
  )
}
abline(v = 0, lty = 2)
legend(
  "topleft",
  legend = colnames(PC2),
  pch = 16,
  col = c("black", rainbow(ncol(PC2))),
  pt.cex = c(2, rep(1, ncol(PC2))),
  bty = "n",
  cex = 0.7
)
```
::: 