---
title: "A multivariate approach"
author: Romain Frelat
date: today
format: 
  html:
    code-fold: true
    toc: true
execute:
  echo: true
  warning: false
---

## Objective

The idea is to use make a PCA over *political stringency* indicators to understand what are the main political stringency "axis" i.e. the main differences among communes in political stringency indicators across France?


```{r}
#| echo: false

# Load home made functions
devtools::load_all()
library(ade4)

ref_folder <- here::here("data", "derived-data", "ref")
ind_folder <- here::here("data", "derived-data", "indicators_csv")
out_folder <- here::here("data", "derived-data", "pca_output")

reg <- terra::vect(file.path(ref_folder, "gadm41_FRA_1.json"))
# dep <- terra::vect(file.path(ref_folder, "gadm41_FRA_2.json"))
shp <- terra::vect(file.path(ref_folder, "commune_simple0005_4326.gpkg"))

meta <- readxl::read_xlsx(here::here("progress", "megatrends_metadata.xlsx"))

# ps_crit <- grepl("stringency", meta$Megatrend)
ind_crit <- meta$Indicator == "yes"
ps <- meta[ind_crit, ]

dfI <- read.csv(file.path(ind_folder, "Dataset_commune.csv"))

ind <- dfI[, ps$Name]
```

### Background information

### Variables

```{r}
#| label: metadata
#| echo: false
colC <- c(
  "Name",
  "Description",
  "Year",
  "Unit",
  "Source"
)
ps <- ps[, colC]
knitr::kable(ps)
```


### Remove outliers

```{r} 
# remove NAs and remove Corsica
keep <- complete.cases(ind) & dfI$INSEE_REG != 94
indF <- ind[keep, ]
df <- dfI[keep, ]

iqr_var <- apply(indF, 2, sd)
med_var <- apply(indF, 2, mean)
low_th <- med_var - iqr_var * 5
high_th <- med_var + iqr_var * 5

out <- t(t(indF) < low_th | t(indF) > high_th)
# apply(out, 2, sum)
# table(apply(out, 1, any))
rm_outlier <- apply(out, 1, any)
# table(keep)
indF <- indF[!rm_outlier, ]
df <- df[!rm_outlier, ]

shp <- shp[match(df$INSEE_COM, shp$INSEE_COM)]
```

We decided to remove:  

- municipalities with missing values on the indicators  (N=```r sum(!complete.cases(ind))```)
- Corsica municipalities (N=```r sum(dfI$INSEE_REG == 94)```)
- outliers identified as $mean ± 5 * sd$ (N = ```r sum(rm_outlier)```)


At the end, we have a table made of ```r nrow(indF)``` municipalities and ```r ncol(indF)``` variables.

::: {.panel-tabset}

##  no transformation

```{r}
par(mar = c(14, 4, 1, 1))
boxplot(
  scale(indF),
  las = 2,
  main = "no outliers"
)
```

##  square root

```{r}
par(mar = c(14, 4, 1, 1))
boxplot(
  scale(apply(indF, 2, sqrt)),
  las = 2,
  main = "sqrt - no outliers"
)
```

##  log

```{r}
par(mar = c(14, 4, 1, 1))
boxplot(
  scale(apply(indF, 2, log1p)),
  las = 2,
  main = "log - no outliers"
)
```

## sqrt with outlier

```{r}
par(mar = c(14, 4, 1, 1))
boxplot(
  scale(apply(ind, 2, sqrt)),
  las = 2,
  main = "sqrt - with outliers",
)
```


::: 



## Correlation

```{r}
#| fig-width: 8
#| fig-height: 8

# names(ind) <- sapply(strsplit(names(ind), "_"), function(x) {
#   paste(x[1], x[2], sep = "\n")
# })
# pairs(ind, lower.panel = panel.smooth.max, upper.panel = panel.cor)
corrplot::corrplot(
  cor(indF, method = "spearman"),
  method = 'number',
  type = 'upper',
  diag = FALSE
)
```


## Principal Component Analysis

```{r}
pca1 <- ade4::dudi.pca(
  indF,
  scannf = FALSE,
  nf = 4
)

barplot(
  pca1$eig / sum(pca1$eig) * 100,
  ylab = "% variance explained",
  col = c(rep("black", 3), rep("grey", ncol(indF) - 3))
)
```

```{r}
#| echo: false
lengthname <- nchar(row.names(pca1$co))
shortlength <- ifelse(lengthname > 15, 15, lengthname)
shortname <- substr(row.names(pca1$co), 1, shortlength)
```

::: {.panel-tabset}
## PC1 - PC2

```{r}
ade4::s.class(pca1$li, as.factor(df$INSEE_REG), col = rainbow(13))
ade4::s.arrow(pca1$co * 5, add.plot = TRUE)
mtext(
  text = paste0("PC1 - ", round(pca1$eig[1] / sum(pca1$eig) * 100), "%"),
  side = 1,
  line = 3
)
mtext(
  text = paste0("PC2 - ", round(pca1$eig[2] / sum(pca1$eig) * 100), "%"),
  side = 2,
  line = 3
)
```

## PC1 - PC3

```{r}
ade4::s.class(
  pca1$li,
  as.factor(df$INSEE_REG),
  col = rainbow(13),
  xax = 1,
  yax = 3
)
ade4::s.arrow(pca1$co * 5, add.plot = TRUE, xax = 1, yax = 3)
mtext(
  text = paste0("PC1 - ", round(pca1$eig[1] / sum(pca1$eig) * 100), "%"),
  side = 1,
  line = 3
)
mtext(
  text = paste0("PC3 - ", round(pca1$eig[3] / sum(pca1$eig) * 100), "%"),
  side = 2,
  line = 3
)
```

:::

## Maps

::: {.panel-tabset}
## PC1
```{r}
#| fig-width: 8
#| fig-height: 8
minmax1 <- max(abs(pca1$li[, 1]))
shp$PC1 <- pca1$li[, 1]
# PC1
terra::plot(
  shp,
  "PC1",
  main = "PC1",
  border = NA,
  breaks = seq(-minmax1, minmax1, length.out = 7)
)
terra::plot(reg, add = TRUE, lwd = 2)


od1 <- order(pca1$co[, 1])
load1 <- paste(shortname[od1], round(pca1$co[od1, 1], 2), sep = ":")
legend(8.3, 48, load1, xpd = NA, bty = "n", cex = 0.6)
```

## PC2

```{r}
#| fig-width: 8
#| fig-height: 8
minmax2 <- max(abs(pca1$li[, 2]))
shp$PC2 <- pca1$li[, 2]
# PC2
terra::plot(
  shp,
  "PC2",
  main = "PC2",
  border = NA,
  breaks = seq(-minmax2, minmax2, length.out = 7)
)
terra::plot(reg, add = TRUE, lwd = 2)

od2 <- order(pca1$co[, 2])
load2 <- paste(shortname[od2], round(pca1$co[od2, 2], 2), sep = ":")
legend(8.3, 48, load2, xpd = NA, bty = "n", cex = 0.6)
```

## PC3

```{r}
#| fig-width: 8
#| fig-height: 8
minmax3 <- max(abs(pca1$li[, 3]))
shp$PC3 <- pca1$li[, 3]
# PC2
terra::plot(
  shp,
  "PC3",
  main = "PC3",
  border = NA,
  breaks = seq(-minmax3, minmax3, length.out = 7)
)
terra::plot(reg, add = TRUE, lwd = 2)

od3 <- order(pca1$co[, 3])
load3 <- paste(shortname[od3], round(pca1$co[od3, 3], 2), sep = ":")
legend(8.3, 48, load3, xpd = NA, bty = "n", cex = 0.6)
```

::: 


## Loadings


::: {.panel-tabset}
## PC1

```{r}
ord1 <- order(pca1$co[, 1])

dotchart(
  pca1$co[ord1, 1],
  pch = 16,
  pt.cex = 1.5,
  labels = row.names(pca1$co)[ord1],
  main = "PC1"
)

abline(v = 0, lty = 2)
```

## PC2

```{r}
ord2 <- order(pca1$co[, 2])

dotchart(
  pca1$co[ord2, 2],
  pch = 16,
  pt.cex = 1.5,
  labels = row.names(pca1$co)[ord2],
  main = "PC2"
)

abline(v = 0, lty = 2)
```

## PC3

```{r}
ord3 <- order(pca1$co[, 3])

dotchart(
  pca1$co[ord3, 3],
  pch = 16,
  pt.cex = 1.5,
  labels = row.names(pca1$co)[ord3],
  main = "PC3"
)

abline(v = 0, lty = 2)
```

::: 

```{r}
knitr::kable(round(pca1$co[, 1:3], 2), row.names = TRUE)
```

## Merged map

::: {.panel-tabset}
## safe scale
```{r}
#| fig-width: 8
#| fig-height: 8
minmax <- max(abs(pca1$li[, 1:3]))
pp <- (((pca1$li[, 1:3] / minmax) + 1) / 2)

colrgb <- rgb(pp[, 1], pp[, 2], pp[, 3], alpha = 1)

terra::plot(
  shp,
  col = colrgb,
  main = "RGB",
  border = NA
)
```

```{r}
# Legend
Ternary::TernaryPlot(
  alab = "PC1 \u2192",
  blab = "\u2190 PC2",
  clab = "PC3 \u2192",
  lab.col = c("red", "darkgreen", "blue"),
  main = "RGB", # Title
  point = "right",
  lab.cex = 0.8,
  grid.minor.lines = 0,
  grid.lty = "solid",
  col = rgb(0.9, 0.9, 0.9),
  grid.col = "white",
  axis.col = rgb(0.6, 0.6, 0.6),
  ticks.col = rgb(0.6, 0.6, 0.6),
  axis.rotate = FALSE,
  padding = 0.08,
  axis.labels = round(seq(-minmax, minmax, length.out = 11), 1)
)

# Colour the background:
cols <- Ternary::TernaryPointValues(rgb)
Ternary::ColourTernary(cols, spectrum = NULL)
# Add the data to the plot
Ternary::AddToTernary(graphics::points, pp, pch = ".")
```


## stretch scale
```{r}
#| fig-width: 8
#| fig-height: 8
scale01 <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}
pp <- apply(pca1$li[, 1:3], 2, scale01)

colrgb <- rgb(pp[, 1], pp[, 2], pp[, 3], alpha = 1)

terra::plot(
  shp,
  col = colrgb,
  main = "RGB",
  border = NA
)
```

```{r}
# Legend
Ternary::TernaryPlot(
  alab = "PC1 \u2192",
  blab = "\u2190 PC2",
  clab = "PC3 \u2192",
  lab.col = c("red", "darkgreen", "blue"),
  main = "RGB", # Title
  point = "right",
  lab.cex = 0.8,
  grid.minor.lines = 0,
  grid.lty = "solid",
  col = rgb(0.9, 0.9, 0.9),
  grid.col = "white",
  axis.col = rgb(0.6, 0.6, 0.6),
  ticks.col = rgb(0.6, 0.6, 0.6),
  axis.rotate = FALSE,
  padding = 0.08,
  axis.labels = c("low", rep("", 9), "high")
)

# Colour the background:
cols <- Ternary::TernaryPointValues(rgb)
Ternary::ColourTernary(cols, spectrum = NULL)
# Add the data to the plot
Ternary::AddToTernary(graphics::points, pp, pch = ".")
```

:::

```{r}
# export
out <- cbind(
  df[, c("NOM_M", "INSEE_COM", "INSEE_DEP", "INSEE_REG")],
  indF,
  pca1$li[, 1:3]
)

write.csv(
  out,
  file = file.path(out_folder, "PCA_commune.csv"),
  row.names = FALSE
)

write.csv(
  pca1$co[, 1:3],
  file = file.path(out_folder, "PCA_loadings.csv"),
)
```